### 2016-04-17 writing/latex/wilf.tex
Wilf-equivs

For each of these patterns there is precisely one occurrence
in any permutation containing the pattern. Now consider
the points in the free box in each case. Each of these regions
must contain an avoider of \(\perm{2,3,1}\) that is of length
\(n-2)\). Therefore these classes are all Wilf-equivalent
and the number of length \(n\) avoiders is
\begin{equation*}
    K_n = C_n - C_{n-2}
\end{equation*}
for \(n \ge 2\) where \(C_n\) is the \(n\)th Catalan number, the number of
\(\perm{2,3,1}\) avoiders of length \(n\).

@@

\subsection{}
The containers of the following patterns can only
have exactly one occurrence.
\begin{equation*}
    \pattern{}{1,2}{0/0,0/1,0/2,1/1,2/0,2/1,2/2}
    \pattern{}{2,1}{0/0,0/1,1/0,1/1,1/2,2/0,2/2}
    \pattern{}{1,2}{0/0,0/1,1/1,1/2,2/0,2/1,2/2}
    \pattern{}{1,2}{0/1,0/2,1/0,1/1,2/0,2/1,2/2}
    \pattern{}{2,1}{0/0,0/1,0/2,1/0,1/2,2/0,2/1}
    \pattern{}{1,2}{0/0,0/1,1/0,1/2,2/0,2/1,2/2}
\end{equation*}
Once again we consider the free boxes, obviously for every pattern
except the first the two boxes are independent, and one contains
any avoider of \(\perm{2,3,1}\) and the other must contain a
decreasing sequence by \LemmaRef{lem:incdec}.
Let us consider the first pattern on it's own. In order to
avoid \(\perm{2,3,1}\) across the free boxes we can add some additional
restrictions
\begin{equation*}
    \raisebox{0.6ex}{
    \begin{tikzpicture}[scale=1, baseline=(current bounding box.center)]
        \fill[pattern=north east lines, pattern color=black!75] (1.5, 0) rectangle +(0.5,1);
        \fill[pattern=north east lines, pattern color=black!75] (1.0, 2) rectangle +(0.5,1);
        \draw (1.5,0) -- (1.5,1)
              (1.5,2) -- (1.5,3);
        \modpattern[2]{}{1,2}{0/0,0/1,0/2,1/1,2/0,2/1,2/2}
    \end{tikzpicture}
    }
\end{equation*}
Now we can see that the top free box must contain a decreasing sequence,
and the bottom must contain an avoider of \(\perm{2,3,1}\) and these
two do not interact in any manner. The containers of this pattern are counted
the same as the other patterns, and due to this they are Wilf-equivalent.

\subsection{}
\subsection{}
The classes containing the following patterns are Wilf-equivalent
\begin{equation*}
    \pattern{}{1,2}{} \text{ and }
    \pattern{}{2,1}{}
\end{equation*}
This is true since the only avoiders of these patterns are the decreasing
sequence and the increasing sequence respectively, and both of these
avoid \(\perm{2,3,1}\) in all cases. There is therefore \(1\) avoider at
every length.

\subsection{}
\subsection{}
\begin{equation*}
    p_1 = \pattern{}{1,2}{1/0,1/1,1/2,2/0,2/1} \text{ and }
    p_2 = \pattern{}{2,1}{0/1,1/0,1/1,1/2,2/1}
\end{equation*}
Let \(\mathcal{L}_1\) be the set of avoiders of \(p_1\), by structural decomposition
around the leftmost point we have

>>>>

For each of these patterns there is precisely one occurrence
in any permutation containing the pattern. Now consider
the points in the free box in each case. Each of these regions
must contain an avoider of \(\perm{2,3,1}\) that is of length
\(n-2\). Therefore these classes are all Wilf-equivalent
and the number of length \(n\) avoiders is
\begin{equation*}
    K_n = C_n - C_{n-2}
\end{equation*}
for \(n \ge 2\) where \(C_n\) is the \(n\)th Catalan number, the number of
\(\perm{2,3,1}\) avoiders of length \(n\).

@@

\subsection{}
The containers of the following patterns can only
have exactly one occurrence.
\begin{equation*}
    \pattern{}{1,2}{0/0,0/1,0/2,1/1,2/0,2/1,2/2}
    \pattern{}{2,1}{0/0,0/1,1/0,1/1,1/2,2/0,2/2}
    \pattern{}{1,2}{0/0,0/1,1/1,1/2,2/0,2/1,2/2}
    \pattern{}{1,2}{0/1,0/2,1/0,1/1,2/0,2/1,2/2}
    \pattern{}{2,1}{0/0,0/1,0/2,1/0,1/2,2/0,2/1}
    \pattern{}{1,2}{0/0,0/1,1/0,1/2,2/0,2/1,2/2}
\end{equation*}
Once again we consider the free boxes, obviously for every pattern
except the first the two boxes are independent, and one contains
any avoider of \(\perm{2,3,1}\) and the other must contain a
decreasing sequence by \LemmaRef{lem:incdec}.
Let us consider the first pattern on it's own. In order to
avoid \(\perm{2,3,1}\) across the free boxes we can add some additional
restrictions
\begin{equation}
    \raisebox{0.6ex}{
    \begin{tikzpicture}[scale=1, baseline=(current bounding box.center)]
        \fill[pattern=north east lines, pattern color=black!75] (1.5, 0) rectangle +(0.5,1);
        \fill[pattern=north east lines, pattern color=black!75] (1.0, 2) rectangle +(0.5,1);
        \draw (1.5,0) -- (1.5,1)
              (1.5,2) -- (1.5,3);
        \modpattern[2]{}{1,2}{0/0,0/1,0/2,1/1,2/0,2/1,2/2}
    \end{tikzpicture}
    }
    \label{eqn:splitboxes}
\end{equation}
Now we can see that the top free box must contain a decreasing sequence,
and the bottom must contain an avoider of \(\perm{2,3,1}\) and these
two do not interact in any manner. The containers of this pattern are counted
the same as the other patterns, and due to this they are Wilf-equivalent.

\subsection{}
The classes containing the following patterns are Wilf-equivalent
\begin{equation*}
    \begin{aligned}
        m_1 = \pattern{}{1,2}{0/0,0/1,1/1,1/2,2/0,2/2} &\text{ and }
        m_2 = \pattern{}{1,2}{0/0,0/1,1/0,2/0,2/1,2/2} \\
        \text{ and }
        m_3 = \pattern{}{1,2}{0/2,1/0,1/1,2/0,2/1,2/2} &\text{ and }
        m_4 = \pattern{}{1,2}{0/1,1/0,1/1,2/0,2/1,2/2}
    \end{aligned}
\end{equation*}

First consider the decomposition of avoiders of \(m_1\) in \(\av{\perm{2,3,1}}\)
around the maximum.
\begin{equation*}
    \mathcal{L}_1 =
    \varepsilon \sqcup
    \decompmax{0/1,1/0}{\scriptstyle \mathcal{G}_1\setminus\varepsilon}{}{}{\mathcal{C}}
    \sqcup
    \decompmax{0/0,0/1,1/0}{}{}{}{\mathcal{L}_1}
\end{equation*}

Now we decompose the avoiders of \(m_2\) around the leftmost point
\begin{equation*}
    \mathcal{L}_2 =
    \varepsilon \sqcup
    \decompleft{0/1,1/0}{\mathcal{C}\setminus\varepsilon}{}{}{\mathcal{C}\setminus\varepsilon}
    \sqcup
    \decompleft{0/1,1/0,1/1}{\mathcal{L}_2}{}{}{}
\end{equation*}
This gives us two generating functions satisfying the following pair of equations
\begin{align}
    L_1(x) &= 1 + xC(x)(G(x)-1) +xL_1(x) \label{eqn:genL1}\\
    \text{ and }L_2(x) &= 1 + x(C(x)-1)^2 + xL_2(x) \label{eqn:genL2}
\end{align}
In order for these two functions to give the same value it is necessary to show that
\eqref{eqn:genL1} and \eqref{eqn:genL2} are equal, this occurs if and only if
\(C(x)(G(x)-1) = (C(x)-1)^2\).
\begin{align*}
        (C(x)-1)^2 &= C(x)(G(x)-1)\\
        \shortintertext{Using the definition of \(G\) and \(C\)}
        x^2C^4(x) &=  (1+xC(x))(xC^2(x)-xC(x))\\
        xC^3(x)&=C(x)-1+xC^3(x)-xC^2(x)\\
        0 &= C(x) - 1 -xC^2(x)\\
        C(x) &= 1 +xC^2(x)
\end{align*}
The final line is always satisfied, and therefore the two generating functions
are equal.

Now we look at the other patterns. In particular note that any container of these
patterns can contain the pattern precisely once, \(m_2\) specifies the
minimum and last point, \(m_3\) and \(m_4\) both use the last point and the
previous block bottom (in the set partition context).

Looking at the structure of a container of \(m_3\) in \(\av{\perm{2,3,1}}\)
\begin{equation*}
    \raisebox{0.6ex}{
        \begin{tikzpicture}[scale=1, baseline=(current bounding box.center)]
            \modpattern[2]{}{1,2}{0/2,1/0,1/1,2/0,2/1,2/2}
            \draw (0.5,0.5) node {\(b_1\)};
            \draw (0.5,1.5) node {\(b_2\)};
            \draw (1.5,2.5) node {\(b_3\)};
        \end{tikzpicture}
    }
\end{equation*}
The boxes \(b_2\) and \(b_3\) must contain a decreasing sequence by
\LemmaRef{lem:incdec}. The box labelled \(b_1\) must contain an avoider
of \(\perm{2,3,1}\). However note that the points in this box can have
interaction with any points in box \(b_2\). If there is just one point
in \(b_2\) then any points in \(b_1\) to the left of this point must
be lower than any points to the right of this point. By extension,
if \(b_2\) contains a decreasing sequence with \(k\) points, there
are \(k+1\) non-interacting avoiders of \(\perm{2,3,1}\) in \(b_3\).

Now in \(m_2\) and \(m_4\) containers we can use the same method as in
\eqref{eqn:splitboxes} to separate the two decreasing sequences in the
free boxes in the top row, and the mixing happens in the same manner
as in a container of \(m_3\).
We now have that \(m_2,m_3\) and \(m_4\) have the same number of containers
so are Wilf-equivalent, and that \(m_1\) and \(m_2\) have the same generating
function. So all four classes are Wilf-equivalent.
\subsection{}
The classes containing the following patterns are Wilf-equivalent
\begin{equation*}
    \pattern{}{1,2}{} \text{ and }
    \pattern{}{2,1}{}
\end{equation*}
This is true since the only avoiders of these patterns are the decreasing
sequence and the increasing sequence respectively, and both of these
avoid \(\perm{2,3,1}\) in all cases. There is therefore \(1\) avoider at
every length.

\subsection{}
The patterns
\begin{equation*}
    \pattern{}{1,2}{0/0,0/2,1/0,1/2,2/0,2/2},
    \pattern{}{1,2}{1/1,2/1,1/0,1/2,2/0,2/2},
    \pattern{}{1,2}{0/1,1/0,1/2,2/0,2/2},
    \pattern{}{1,2}{0/1,0/2,1/1,1/2,2/0}, \text{ and }
    \pattern{}{1,2}{0/0,0/1,0/2,1/0,1/1,1/2}
\end{equation*}
are Wilf-equivalent.
If \(\mathcal{M}_1\) is the set of avoiders of
\textpattern{}{1,2}{0/0,0/2,1/0,1/2,2/0,2/2}, then by the structural
decomposition around the maximum we have
\begin{equation*}
    \mathcal{M}_1 = \varepsilon \sqcup
    \decompmax{0/0,0/1,1/0}{}{}{}{\mathcal{C}}
\end{equation*}
This is because \(\mathcal{M}_1\) is the set of permutations
who have their minimum occur after their maximum.

The second pattern occurs if the last element is higher than
the penultimate element. This can only occur if the last element is
in a single block in the set partition context, In order to construct
a avoider of length \(n\) we can take any avoider \(\perm{2,3,1}\)
of length \(n-1\) and insert the new maximum into the last block.
This ensures that the last block is never a singleton. This means
that these permutations are also counted by \(M(x) = 1+xC(x)\).

The avoiders of the third pattern can be decomposed by the maximum
to give
\begin{equation*}
    \mathcal{M}_3 = \varepsilon \sqcup
    \decompmax{0/1,1/0}{\mathcal{F}}{}{}{\mathcal{M}_3}
\end{equation*}
The generating function derived satisfies \(M_3(X) = 1 + xF(x)M_3(x)\).
The fourth pattern can be decomposed around the maximum in a similar
manner.
\begin{equation*}
    \mathcal{M}_4 = \varepsilon \sqcup
    \decompmax{0/1,1/0}{\mathcal{M}_4}{}{}{\mathcal{F}}
\end{equation*}

Finally considering the last pattern, the only way we can construct an
avoider is to take any \(\perm{2,3,1}\) avoider and add a new minimum
at the start of the permutation. Adding a new leftmost point in any other
position would either create an occurrence of \(\perm{2,3,1}\) or the pattern.
Therefore this is also counted by \(M(x) = 1+xC(x)\).
\subsection{}
\begin{equation*}
    p_1 = \pattern{}{1,2}{1/0,1/1,1/2,2/0,2/1} \text{ and }
    p_2 = \pattern{}{2,1}{0/1,1/0,1/1,1/2,2/1}
\end{equation*}
Let \(\mathcal{N}_1\) be the set of avoiders of \(p_1\), by structural decomposition
around the leftmost point we have
\begin{equation*}
    \mathcal{N}_1 = \varepsilon \sqcup
    \decompleft{0/1,1/0}{\mathcal{N}_1}{}{}{\mathcal{N}^\prime_1}
\end{equation*}
Where \(\mathcal{N}^\prime_1\) is a permutation avoiding \(\perm{2,3,1}, m_1\)
and \(\textpattern{}{1}{0/0,0/1,1/1}\)
Now consider the decomposition of a permutation in \(\mathcal{N}^\prime_1\) it can
once again be decomposed around the leftmost point
\begin{equation*}
    \mathcal{N}^\prime_1 = \varepsilon \sqcup
    \decompleft{0/1,1/0}{\scriptstyle \mathcal{N}_1\setminus\varepsilon}{}{}{\mathcal{N}^\prime_1}
\end{equation*}
This is a complete decomposition of avoiders of \(p_1\). Now we
look at an avoider of \(p_2\), this time decomposition is around the maximum
\begin{equation*}
    \mathcal{N}_2 = \varepsilon \sqcup
    \decompleft{0/1,1/0}{\mathcal{N}_2}{}{}{\mathcal{N}^\prime_2}
\end{equation*}

Again we use the same method of decomposition of a permutation in \(\mathcal{N}^\prime_2\)
\begin{equation*}
    \mathcal{N}^\prime_2 = \varepsilon \sqcup
    \decompleft{0/1,1/0}{\scriptstyle \mathcal{N}_2\setminus\varepsilon}{}{}{\mathcal{N}^\prime_2}
\end{equation*}

This gives us a generating function \(N(x)\) satisfying
\begin{align}
    N(x) &= 1 + xN(x)N^\prime(x) \label{eqn:Ngen}\\
    N(x) &= 1 + x(N(x)-1)N^\prime(x)\label{eqn:NprimeGen}
\end{align}

Solving \EquationRef{eqn:NprimeGen} for \(N^\prime(x)\) and substituting into
\EquationRef{eqn:Ngen} gives us the fact that a the generating function for
\(N(x)\) satisfies
\begin{equation}
    N(x) = xN^2(x) - x(N(x) - 1) + 1
\end{equation}
